Continuous Deductive Reasoning over Streamed RDF Data

### 2015-07-10 content/contribution/CoDeR.tex

	% Stream-based Data Model
	\begin{nestedsection}{Data Model}{model: data}
		The data model of CoDeR is composed of two classes of streams: \emph{input}/\emph{output} streams and \emph{inter-operator} streams.
		In both cases, streams are multi-source multi-sink logical entities.

@@

		The interval ${(i_e,i_n]}$ for any graph or triple instance is the the period of time for which the data is \emph{valid} \citep{SemanticStreamingManagement,sparkwave} according to the ${\Pi^C = \langle R, F, S, r \rangle}$ expressed.
		For graphs on the input stream, $i_e$ is the time at which a given graph was asserted, and ${i_n = i_e + r}$ is the time at which it will leave the semantic sliding window over $S$ of $\Pi^C$.
		For graphs on the output stream, $i_e$ is the latest time at which any part of the justification of a given graph was asserted, and $i_n$ is the earliest time at which any part of its justification will leave the sliding window.
		\reffig{intersected-intervals} provides an example of this for some ${\Pi^C = \langle \{c\,\text{:-}\,a,b\,.\},\{\},(\{a\}\text{:}1,\{b\}\text{:}2),3\}}$ over instants ${i = 1 \dots 4}$.
		\begin{figure}
			\centering
			\includegraphics[width=0.45\textwidth]{intersected-intervals}
			\caption{Derivation of valid-time for entailed fact instance from those of its justifying fact instances.}
			\labelfig{intersected-intervals}
		\end{figure}
		\begin{align*}
			CDP_{1} & = \{c\,\text{:-}\,a,b.\} \cup \{ a\text{:}1. \} & \vDash & \{ c\,\text{:-}\,a,b. \, a\text{:}1. \} \\
			CDP_{2} & = \{c\,\text{:-}\,a,b.\} \cup \{ a\text{:}1. \, b\text{:}2. \} & \vDash & \{ c\,\text{:-}\,a,b. \, a\text{:}1. \, b\text{:}2. \, c\text{:}2. \} \\
			CDP_{3} & = \{c\,\text{:-}\,a,b.\} \cup \{ a\text{:}1. \, b\text{:}2. \} & \vDash & \{ c\,\text{:-}\,a,b. \, a\text{:}1. \, b\text{:}2. \, c\text{:}2. \} \\
			CDP_{4} & = \{c\,\text{:-}\,a,b.\} \cup \{ b\text{:}2. \} & \vDash & \{ c\,\text{:-}\,a,b. \, b\text{:}2. \}
		\end{align*}
		It is clear that a CoDeR output stream expresses the semantic stream $P$ entailed by some $\Pi^C$.
		However, as the semantic stream $N$ is defined as the sequence of entailments of some $\Pi^C$ ordered by the instant $i_{n}$ at which they \emph{ceased} to be justified by $W^{S,r}_{i_n}$, the output stream of a CoDeR-based system also expresses the stream $N$.
		As such, the annotation of streamed RDF graphs with the interval $(i_e,i_n]$ gives the streams of CoDeR a dual semantics with regard to Continuous Datalog.

@@

	% Minimal Algebra
	\begin{nestedsection}{Minimal Algebra}{model: algebra}
		The minimal algebra of CoDeR is based on that of Positive Datalog, that being ${\{\bot, \text{:-}, \wedge\}}$.
		Of these, implication $\text{:-}$ and conjunction $\wedge$ may be expressed as stream-to-stream operators, the former being extended to apply to instances as $\text{:-}_{head}$ and the latter being the window-join $\Join$ \citep{niagaraCQ}, in order to support forward-chaining processing of fixed rules;
		$\bot$ is not an operator, but the special class of values that should not exist, represented in CoDeR by the ${rif\text{:}error()}$ class from RIF \citep{w3crif}.
		% Further, in order to support both fixed-plan pattern-matching algorithms such as the Rete algorithm \citep{forgy79} and adaptive algorithms such as the Eddies algorithm \citep{eddies} and its derivatives \citep{CACQ,TCQ}, CoDeR replaces the window-join operator with the State Module of Eddies, algebraically represented by ${{}_b\Join_p}$.
		Further, in order to support both fixed-plan pattern-matching algorithms such as the Rete algorithm \citep{forgy79} and adaptive algorithms such as the Eddies algorithm \citep{eddies} and its derivatives \citep{CACQ,TCQ}, CoDeR replaces the window-join operator with the State Module of Eddies, algebraically represented by $\rstreamjoin$.
		% In addition to $\text{:-}_{head}$ and ${{}_b\Join_p}$, CoDeR utilises a pair of stream-filter operators, $\sigma_{triple}$ and $\sigma_{external}$, for identifying \emph{basic patterns} as defined in SPARQL \citep{w3csparql} and expressing the datatype functions of RIF-Core \citep{w3crifcore}, respectively.
		In addition to $\text{:-}_{head}$ and $\rstreamjoin$, CoDeR utilises a pair of stream-filter operators, $\sigma_{triple}$ and $\sigma_{external}$, for identifying \emph{basic patterns} as defined in SPARQL \citep{w3csparql} and expressing the datatype functions of RIF-Core \citep{w3crifcore}, respectively.
		Finally, as basic patterns match individual triples, the CoDeR algebra includes a graph-to-triples operator ${{}_g\pi_t}$ to translate streams $S$ from the streams of graphs recommended by the RSP community\footnote{\url{https://github.com/streamreasoning/RSP-QL/blob/master/Semantics.md}} to the streams of triples on which $\sigma_{triple}$ operates.
		% In summary, the minimal algebra of CoDeR consists of the set of operators ${\{\text{:-}_{head}, {}_b{\Join_p}, \sigma_{triple}, \sigma_{external}, {}_g{\pi_t}\}}$.
		In summary, the minimal algebra of CoDeR consists of the set of operators ${\{\text{:-}_{head}, \rstreamjoin, \sigma_{triple}, \sigma_{external}, {}_g{\pi_t}\}}$.

		As all operators are stream-to-stream in nature, they each take one or two streams as input and produce a single stream (though they may contribute to any number of union streams).
		As such, the notation for applying an operator to a stream is also the definition of a stream, and may, therefore, be used as input to another operator.
		For example, \refeqn{basic-example-plan} is a plan expressed in the CoDeR algebra for the ${CDP = \{ \{b\,\text{:-}\,a.\}, S, R \}}$, where $a$ and $b$ are RDF triples.
		\begin{equation}\labeleqn{basic-example-plan}
			\text{:-}_b \left( \sigma_a \left( {}_g{\pi_t}\,S \right) \right)
		\end{equation}

@@

			\item[$\sigma_{triple}\,S$] \hfill \\

@@

				The triple-select operator identifies instances of triples in a stream $S$ that match some \emph{basic pattern} \citep{w3csparql}.

@@

				It takes an inter-operator stream $S$ of triple instances as input and produces an inter-operator stream of triple instances, where the produced stream contains a subset of those instances in the input stream, maintaining the ordering of those instances.

@@

			\item[$\sigma_{external}\,S$] \hfill \\

@@

				It filters elements of its input stream $S$ based not on the semantic structure of the data but directly on the values of variables, which are consistent throughout any given graph pattern;

@@

				As such, it takes an inter-operator stream $S$ of graph instances as input and produces an inter-operator stream of graph instances, where the produced stream contains a subset of those instances in the input stream, maintaining the ordering of those instances.

@@

			\item[${{}_g{\pi_t}\,S}$] \hfill \\

@@

				It takes a heterogeneous (\emph{input/output}) stream $S$ of RDF graph instances and produces a homogeneous (\emph{inter-operator}) stream of RDF triple instances.

@@

			\item[$\text{:-}_{head}\,S$] \hfill \\

@@

				Instance implication is a one-pass operator that takes as input an inter-operator stream $S$ whose graph pattern is the body of one of the axiomatic rules $A$ of the expressed ${CDP}$, and produces an inter-operator stream whose graph pattern is the head of that axiomatic rule.

@@

			\item[${S_1\,\rstreamjoin\,S_2}$] \hfill \\

@@

				The implementations of SteMs in the literature show a great variety of behaviour regarding window maintenance \citep{SteMs}, in the most extreme case ceasing to be a wholly stream-to-stream operator and supporting the incorporation of ``static'' data into stream processing systems.
				In the minimal algebra of CoDeR, however, SteMs simply support the building of a \emph{valid window} (as defined in \refsec{semantics}) from its first inter-operator stream $S_1$, and the probing of that window with data from its second inter-operator stream $S_2$ to produce an inter-operator stream of the homogeneously joined graph instances.
				Furthermore, the nature of probing in CoDeR is strictly limited to that of the natural join between the variables in the patterns for streams $S_1$ and $S_2$.
				Finally, in contrast to the direct inheritance of valid-times in unary operators of CoDeR, the valid-time interval assigned to each graph produced by a SteM is the intersection of the valid-times of the two joined graphs.
		\end{description}

		% As the SteM ${{}_b{\Join_p}}$ is the only CoDeR operator that maintains any form of window over its input, and SteMs maintain exclusively \emph{valid windows}, no data will persist within a CoDeR-based system when it is no longer valid according to its annotations.
		As the SteM $\rstreamjoin$ is the only CoDeR operator that maintains any form of window over its input, and SteMs maintain exclusively \emph{valid windows}, no data will persist within a CoDeR-based system when it is no longer valid according to its annotations.
		Furthermore, the valid-time interval of any instance is either directly inherited from the interval of the input instance, in the case of one-pass operators, or derived from the intersections of the intervals of both contributing instances in the sole two-pass operator, the SteM;
		as such, it is guaranteed that all instance annotations accurately describe the intervals for which the given instance is justified by the contents of the semantic sliding window of range $R$ over the system input streams $S$ defined in a given ${CDP = \{A,S,R\}}$.
		It follows that no facts may persist in a CoDeR based system if it is no longer valid, and thus only entailments that are valid at a given moment may be produced in that moment, thereby guaranteeing correct results according to the semantics of Continuous Datalog.
		It also follows that, given a system with a processing plan that is expressible in the CoDeR algebra and accurately represents a set of rules $A$ of a ${CDP}$, all facts that are logically entailed by that ${CDP}$ at a given instant will be persisted within that system at that instant,
		i.e. such a CoDeR-based system is guaranteed to produce the complete set of results according to the semantics of Continuous Datalog.
	\end{nestedsection}

%	\subimport{conclusions/}{future-work.tex}
\end{nestedsection}

>>>>

	% Stream-based Data Model
	\begin{nestedsection}{Data Model}{model: data}
		The data model of CoDeR is composed of multiple-source multiple-sink streams.
		These streams are divided into two classes: \emph{input}/\emph{output} streams and \emph{inter-operator} streams.

@@

		The interval ${(i_e,i_n]}$ for any graph or triple instance is the the period of time for which the data is \emph{valid} \citep{SemanticStreamingManagement,sparkwave} according to the ${\Pi^C = \langle R, F, S, r \rangle}$ expressed.
		For graphs on the input stream, $i_e$ is the time at which a given graph was asserted, and ${i_n = i_e + r}$ is the time at which it will leave the semantic sliding window over $S$ of $\Pi^C$.
		For graphs on the output stream, $i_e$ is the latest time at which any part of the justification of a given graph was asserted, and $i_n$ is the earliest time at which any part of its justification will leave the sliding window.
		\reffig{intersected-intervals} provides an example of this for some ${\Pi^C = \langle \{c\,\text{:-}\,a,b\,.\},\{\},(\{a\}|1,\{b\}|2),3\}}$ over instants ${i = 1 \dots 4}$.
		\begin{figure}
			\centering
			\includegraphics[width=0.45\textwidth]{intersected-intervals}
			\caption{Derivation of valid-time for entailed fact instance from those of its justifying fact instances.}
			\labelfig{intersected-intervals}
		\end{figure}
		\begin{align*}
			\Pi^C_{1} & = \langle \{c\,\text{:-}\,a,b.\}, \{ a|1. \} \rangle & \vDash & \{ \} \\
			\Pi^C_{2} & = \langle \{c\,\text{:-}\,a,b.\}, \{ a|1. \, b|2. \} \rangle & \vDash & \{ c|2. \} \\
			\Pi^C_{3} & = \langle \{c\,\text{:-}\,a,b.\}, \{ a|1. \, b|2. \} \rangle & \vDash & \{ c|2. \} \\
			\Pi^C_{4} & = \langle \{c\,\text{:-}\,a,b.\}, \{ b|2. \} & \vDash & \{ \}
		\end{align*}
		It is clear that a CoDeR output stream expresses the semantic stream $P$ entailed by some $\Pi^C$.
		However, as the semantic stream $N$ is defined as the sequence of entailments of some $\Pi^C$ ordered by the instant $i_{n}$ at which they \emph{ceased} to be justified by $W^{S,r}_{i_n}$, the output stream of a CoDeR-based system also expresses the stream $N$.
		As such, the annotation of streamed RDF graphs with the interval $(i_e,i_n]$ gives the streams of CoDeR a dual semantics with regard to Continuous Datalog.

@@

	% Minimal Algebra
	\begin{nestedsection}{Minimal Algebra}{model: algebra}
		The minimal algebra of CoDeR is based on that of Positive Datalog, that being ${\{\bot, \text{:-}, \wedge\}}$.
		Of these, implication $\text{:-}$ and conjunction $\wedge$ may be expressed as stream-to-stream operators, the former being extended to apply to instances as $\text{:-}_{head}$ and the latter being the window-join $\Join$ \citep{niagaraCQ}, in order to support forward-chaining processing of fixed rules;
		$\bot$ is not an operator, but the special class of values that should not exist, represented in CoDeR by the ${rif\text{:}error()}$ class from RIF \citep{w3crif}.
		% Further, in order to support both fixed-plan pattern-matching algorithms such as the Rete algorithm \citep{forgy79} and adaptive algorithms such as the Eddies algorithm \citep{eddies} and its derivatives \citep{CACQ,TCQ}, CoDeR replaces the window-join operator with the State Module of Eddies, algebraically represented by ${{}_b\Join_p}$.
		Further, in order to support both fixed-plan pattern-matching algorithms such as the Rete algorithm \citep{forgy79} and adaptive algorithms such as the Eddies algorithm \citep{eddies} and its derivatives \citep{CACQ,TCQ}, CoDeR replaces the window-join operator with the State Module of Eddies, which we represent algebraically by $\rstreamjoin$.
		% In addition to $\text{:-}_{head}$ and ${{}_b\Join_p}$, CoDeR utilises a pair of stream-filter operators, $\sigma_{triple}$ and $\sigma_{external}$, for identifying \emph{basic patterns} as defined in SPARQL \citep{w3csparql} and expressing the datatype functions of RIF-Core \citep{w3crifcore}, respectively.
		In addition to $\text{:-}_{head}$ and $\rstreamjoin$, CoDeR utilises a pair of stream-filter operators, $\sigma_{triple}$ and $\sigma_{external}$, for identifying \emph{basic patterns} as defined in SPARQL \citep{w3csparql} and expressing the datatype functions of RIF-Core \citep{w3crifcore}, respectively.
		Finally, as basic patterns match individual triples and CoDeR's streams are composed of graphs, the CoDeR algebra includes a graph-to-triples operator ${{}_g\pi_t}$ to translate input streams to the streams of triples on which $\sigma_{triple}$ operates.
		% In summary, the minimal algebra of CoDeR consists of the set of operators ${\{\text{:-}_{head}, {}_b{\Join_p}, \sigma_{triple}, \sigma_{external}, {}_g{\pi_t}\}}$.
		In summary, the minimal algebra of CoDeR consists of the set of operators ${\{\text{:-}_{head}, \rstreamjoin, \sigma_{triple}, \sigma_{external}, {}_g{\pi_t}\}}$.

		As all operators are stream-to-stream in nature, they each take one or two streams as input and produce a single stream (though they may contribute to any number of union streams).
		As such, the notation for applying an operator to a stream is also the definition of a stream, and may, therefore, be used as input to another operator.
		For example, \refeqn{basic-example-plan} is a plan expressed in the CoDeR algebra for the ${\Pi^C = \{ \{b\,\text{:-}\,a.\}, \{\}, S, r \}}$, where $a$ and $b$ are RDF triples.
		\begin{equation}\labeleqn{basic-example-plan}
			\text{:-}_b \left( \sigma_a \left( {}_g{\pi_t}\,S \right) \right)
		\end{equation}

@@

			\item[$\sigma_{triple}\,X$] \hfill \\

@@

				The triple-select operator identifies instances of triples in a stream $X$ that match some \emph{basic pattern} \citep{w3csparql}.

@@

				It takes an inter-operator stream $X$ of triple instances as input and produces an inter-operator stream of triple instances, where the produced stream contains a subset of those instances in the input stream, maintaining the ordering of those instances.

@@

			\item[$\sigma_{external}\,X$] \hfill \\

@@

				It filters elements of its input stream $X$ based not on the semantic structure of the data but directly on the values of variables, which are consistent throughout any given graph pattern;

@@

				As such, it takes an inter-operator stream $X$ of graph instances as input and produces an inter-operator stream of graph instances, where the produced stream contains a subset of those instances in the input stream, maintaining the ordering of those instances.

@@

			\item[${{}_g{\pi_t}\,X}$] \hfill \\

@@

				It takes a heterogeneous (\emph{input/output}) stream $X$ of RDF graph instances and produces a homogeneous (\emph{inter-operator}) stream of RDF triple instances.

@@

			\item[$\text{:-}_{head}\,X$] \hfill \\

@@

				Instance implication is a one-pass operator that takes as input an inter-operator stream $X$ whose graph pattern is the body of one of the rules $R$ of the expressed $\Pi^C$, and produces an inter-operator stream whose graph pattern is the head of that axiomatic rule.

@@

			\item[${X\,\rstreamjoin\,Y}$] \hfill \\

@@

				The implementations of SteMs in the literature show a great variety of behaviour regarding window maintenance \citep{SteMs}, in the most extreme case ceasing to be a wholly stream-to-stream operator and supporting the incorporation of ``static'' data into stream processing systems.
				In the minimal algebra of CoDeR, however, SteMs simply support the building of a \emph{valid window} (as defined in \refsec{semantics}) from its first inter-operator stream $X$, and the probing of that window with data from its second inter-operator stream $Y$ to produce an inter-operator stream of the homogeneously joined graph instances.
				Furthermore, the nature of probing in CoDeR is strictly limited to that of the natural join between the variables in the patterns for streams $X$ and $Y$.
				Finally, in contrast to the direct inheritance of valid-times in unary operators of CoDeR, the valid-time interval assigned to each graph produced by a SteM is the intersection of the valid-times of the two joined graphs.
		\end{description}

		% As the SteM ${{}_b{\Join_p}}$ is the only CoDeR operator that maintains any form of window over its input, and SteMs maintain exclusively \emph{valid windows}, no data will persist within a CoDeR-based system when it is no longer valid according to its annotations.
		As the SteM $\rstreamjoin$ is the only CoDeR operator that maintains any form of window over its input, and SteMs maintain exclusively \emph{valid windows}, no data will persist within a CoDeR-based system when it is no longer valid according to its annotations.
		Furthermore, the valid-time interval of any instance is either directly inherited from the interval of the input instance, in the case of one-pass operators, or derived from the intersections of the intervals of both contributing instances in the sole two-pass operator, the SteM;
		as such, it is guaranteed that all instance annotations accurately describe the intervals for which the given instance is justified by the contents of the semantic sliding window of range $r$ over the system input streams $S$ defined in a given ${\Pi^C = \{R,F,S,r\}}$.
		It follows that no facts may persist in a CoDeR based system if it is no longer valid, and thus only entailments that are valid at a given moment may be produced in that moment, thereby guaranteeing correct results according to the semantics of Continuous Datalog.
		It also follows that, given a system with a processing plan that is expressible in the CoDeR algebra and accurately represents a set of rules $R$ of a $\Pi^C$, all fact instances that are logically entailed by that $\Pi^C$ at a given instant will be persisted within that system at that instant,
		i.e. such a CoDeR-based system is guaranteed to produce the complete set of results according to the semantics of Continuous Datalog.
	\end{nestedsection}
\end{nestedsection}
